Do motif roles provide unique information about species risk of extinction?

### 2021-08-13 manuscript/motifs_vs_degree.tex

            A species' motif participation is likely related to its degrees and trophic level since all three are measures of network position.
            To understand how this may have affected the results above, we first fit two linear mixed-effects regressions relating persistence to degree or trophic level, $\pi_{disturbed}$, and their interaction.
            As above, we centered and scaled the predictors.
            We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks. 
            These regressions provide context about how persistence varies with simpler measures of species' roles within networks.

@@

        % \subsubsection*{How does persistence vary with motif positions?}

        %     As one more level of detail, we can consider the frequency with which a species appears in each \emph{position} in each motif. 
        %     Different positions are associated with different numbers of predators and prey, making the similarities and differences between motifs and degree more explicit.
        %     The full set of these positions, across all motifs, is the species' \emph{motif role}~\citep{Stouffer2010,Cirtwill2017}.
        %     As with motif participation, we normalized motif roles by dividing the number of times a species appeared in each position by the total number of times it appeared in all positions~\citep{Baker2015,Cirtwill2017}.

        %     [[Could add a PERMANOVA to show that species' roles overal affect their persistence. Haven't done so yet.]]

        %     To test whether participating in different motif positions was related to a species' probability of persistence, we fit similar models to those for motif participation.
        %     For each position, we fit a linear mixed-effect model relating persistence to the frequency of the position in a species' motif role, disturbance, and their interaction.
        %     To account for differences in persistence across networks of different size and connectance, we included a random effect of the interaction between size and connectance.
        %     All regressions were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.

        % \subsubsection*{How do motif roles vary with other network properties?}

        %     As with network motif profiles and motif participation, species' motif roles may vary with network size and connectance and with the species' in degree and trophic level.
        %     To identify these relationships, we fit two linear regressions for each motif position.
        %     The first related the frequency of the position in species' roles to network size, connectance, and their interaction.
        %     The second related the frequency of the position to in-degree, STL, and their interaction.
        %     All regressions were fit using the R~\citep{R} base function `lm'.

@@

    We are interested in i) how the distribution of motifs in a network (i.e., the network's \emph{motif profile}) affect network persistence, and ii) how a species motif participation affects its persistence. We analyse this for different levels of disturbance to the basal level. To provide context, we also investigate relationships between persistence and classic global (size and connectance) and local (in-degree and trophic level) network properties.
    % The results are organized as follows: 1) How network persistence varies with network motif distribution. 2) How network persistence varies with global structural properties. 3) How network motif distribution varies with global structural properties. 4) How species probability of persistence varies with its motif profile. 5) How species persistence varies with its characteristics such as TL and in-degree. 6) How a species motif profile depend on its characteristics. 7) How a species persistence varies with its motif role (position in respective motif).

@@

        \begin{figure}[h!]
            \centering
            \includegraphics[width=.75\textwidth]{manuscript/figures/motif_proportion_lms.pdf}
            \caption{The proportions of a network's motif profile made up by the four motifs which can appear in acyclic networks vary with the size and connectance of the network. More-connected networks tend to contain relatively more omnivory motifs and fewer apparent competition motifs. The proportions of direct competition and three-species chain motifs were less variable.}
            \label{motif_proportion_lms}
        \end{figure}

@@

        The overall motif profile of a network is related to the average persistence of consumers ($F_{1,2998}$=571, $p$\textless0.001). Average persistence is higher in networks with higher proportions of the chain, direct competition, and apparent competition motifs (Table S3, \emph{Appendix SC}). Conversely, average persistence is lower in networks with higher proportions of the omnivory motif. These trends hold for all levels of disturbance (Fig.~\ref{motif_profile_persistence}, Table S3), \emph{Appendix SC}).% Network persistence varies between networks with different dispersion of motif profiles ($F_{134,2865}$=2.19$\times10^{24}$, $p$=\textless0.001). For example, networks with more variable motif profiles have higher persistence ($\beta$=0.183, $p$\textless0.001).

        \begin{figure}
            \centering
            \includegraphics[width=.5\textwidth]{figures/persistence_motif_profiles.eps}
            \caption{The proportions of the four motifs in a network's motif profile are related to the mean persistence of species in the network. Specifically, persistence decreases as the proportion of omnivory in the network's motif profile increases while persistence increases with the proportions of the other three motifs. Interactions between motif profiles and disturbance were significant but small. Here we show the relationships of probabilities of extinction of basal resources of 0.1 (top set of lines) and 0.5 (bottom set of lines). Relationships are shown for the observed range of proportions for each motif.}      
            \label{fig:motif_profile_persistence}
        \end{figure}

       The effects of motif participation on a species' probability of persistence depends on the level of disturbance. When disturbance is low, species' persistence increases with increasing proportions of the omnivory and direct competition motifs. Conversely, a species' persistence decreases with an increasing proportion of the apparent competition motif. Persistence is not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}).

        At high levels of disturbance, a species' persistence decreases as the proportion of the three-species chain increases, but persistence increases as the proportion of any other motif increases (Figure~\ref{fig:prop_lmer_all}). This means that increased disturbance strengthens the effects of the three-species chain motif and changes the direction of the effect of the apparent competition motif  (Table S4, \emph{Appendix SD}). The effect of omnivory on persistence is still positive but weaker at higher levels of disturbance, and the relationship between persistence and direct competition was largely unaffected by disturbance.

            \begin{figure}[h!]
                \centering
                \includegraphics[width=0.85\textwidth]{figures/prop_lmer_allCS.pdf}
                \caption{The effect of proportion of the role (x-axis) made up by various motifs (columns) on persistence (y-axis). The effect of participating in each motif is based on linear mixed-effect models following Equation~\ref{propreq}, except that models for each  disturbance level are fitted separately. The different colored lines indicate the probability of extinction of basal species, from $\pi_{disturbed} = 0.1$ (top, purple; no disturbance) to $\pi_{disturbed} = 0.5$ (bottom, yellow; high disturbance). Note that omnivory made up a smaller proportion of species' roles than other motifs; lines are plotted over the observed ranges of motif participation.}
                \label{fig:prop_lmer_all}
            \end{figure}

        \subsubsection*{Consistency in effects of motif participation on species persistence}

            % In order to evaluate how consistent the relationships between a species persistence and motif participation is, and how it varies with  level of disturbance and global network structure, we fit linear regressions of species persistence against motif proportions for the species for each network separately. We then analyze the density distribution of the values of the slopes. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here.

            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). However, there is an interaction between disturbance and connectance for the direct competition and omnivory motifs; At low and medium connectances there is little change in the proportion of positive slopes when the proportion of  direct competition motif is high, independent of level of disturbance. At high connectances, however, the proportion of positive slopes is lower and also decreases with increasing disturbance. For the omnivory motif the trend is reversed. Additionally, the omnivory motif stands out having a wider curve when network connectance is low. 
            Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}).

        \begin{figure}[h!]
            \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_C_allS.pdf}
            \caption{Density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all sizes - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. Columns show the result for various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (left) to $\pi_{disturbed} = 0.5$ (right). Rows show various levels of connectance. The dotted, vertical line indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_C}
        \end{figure}

@@

        A species persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix SF}). Increased trophic level is generally associated with decreasing species persistence, although the rate of decline is slower at high disturbance (Table S5, \emph{Appendix SF}).
        In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

        % Anna notes (just for our (mine) understanding): 
        % When looking at the two figures of degree vs persistence and STL vs persistence - my conclusion is that species at lower trophic levels have the highest in-degree. Double check - yes, that’s true.
        % Lower TL are all well at low disturbances, and when disturbance increase, persistence decrease (looking at fig degree vs persistence, right end of the line).
        % Higher TL experience the same thing, although to a slightly smaller extent (looking at the left end of the lines in fig degree vs persistence and right end of lines in fig STL vs persistence). 
        % The slope of the line in fig degree vs persistence is not really relevant, since it's kind of based on the response of different TL level (which makes perfect sense in the BN framework)
        % weird thing though - at high levels of disturbance, lower TL seems more affected than higher TL in the persistence vs degree figure (since low degree belongs to higher TL, high degree to lower TL). This does not correspond with fig persistence vs STL. HOWEVER - not only high TL have a low degree, lower TL do as well, probably affecting the overall result. Only low TL have high degree - but the opposite is not true. Probably explains it

@@

            \begin{figure}
                \centering
                \includegraphics[width=\textwidth]{figures/roles_vs_TL.eps}
                \caption{As with motifs, persistence varied with simpler measures of species roles in networks. Moreover, motif frequencies correlated with in-degree and trophic level. \textbf{A-B)} Persistence increased with increasing in-degree for low levels of disturbance to basal resources but decreased with increasing degree for high levels of disturbance to basal resources.
                Persistence declined with inceasing STL for all levels of disturbance to basal resources.
                \textbf{C-D)} The proportion of omnivory increased with increasing degree while all other proportions decreased. The proportions of omnivory and direct competition decreased with increasing STL while the proportions of apparent competition and three-species chains increased.}
                \label{fig:motifs_vs_TL_and_deg}
            \end{figure}

    % \subsection*{Persistence and species' positioning in motifs}

    %     A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table SX, \emph{Appendix SH}).
    %     In the omnivory motif, persistence increased with the frequency of the bottom position (two predators, no prey within the motif) across all levels of disturbance but most steeply at high levels of disturbance (Fig.~\ref{figs:persistence_position_motifs}).
    %     Persistence increased slightly with the frequency of the middle position (one prey, one predator) at low levels of disturbance but showed little relationship at higher levels of disturbance.
    %     Persistence increased with the frequency of the top position (two prey, no predators within the motif) at low disturbance but decreased at higher levels of disturbance.

    %     In the three-species chain, persistence decreased as the frequency of the top position increased at all levels of disturbance.
    %     Conversely, persistence increased with increasing frequency of the middle position at all levels of disturbance.
    %     Persistence decreased slightly with increasing frequency of the bottom position at low levels of disturbance but increased at high levels of disturbance.

    %     In the apparent competition motif, persistence decreased slightly with increasing frequencies of the bottom position (one predator, no prey within the motif) at low disturbance but increased slightly at high disturbance.
    %     Persistence increased with increasing frequency of the top position (two prey, no predator within the motif) at low disturbance but decreased at high disturbance.

    %     In the direct competition motif, persistence varied little with the frequency of the bottom position (two predators, no prey within the motif) at low disturbance but increased with the frequency of this position at high levels of disturbance.
    %     Persistence increased slightly with the frequency of the top position (one prey, no predators within the motif) at low levels of disturbance and decreased slightly at high levels of disturbance.

    % \begin{figure}[h!]
    %     \includegraphics[width=.95\textwidth]{figures/persistence_positions_Omnivory.eps}
    %     \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the omnivory motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
    %     \label{figs:persistence_position_motifs}
    %     \end{figure}

    %     The frequencies of motif positions in a species' role varied with network properties (\emph{Appendix SI}).Motif roles had stronger relationships with degree and STL than with network size or connectance.

>>>>

            A species' motif participation is likely related to its in-degree and trophic level since all three properties are measures of network position.
            To provide context for the results above, we therefore fit two linear mixed-effects regressions relating persistence to degree or trophic level, $\pi_{disturbed}$, and their interaction.
            As above, we centered and scaled the predictors.
            We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks.

@@

    % % % Things in SI that are not referred to:
    % persistence vs. motif positions
    % motif roles (positions) vs. other network properties
    % persistence vs. % Basal
    % motif profiles vs. % Basal

@@

    We were interested in i) how the distribution of motifs in a network (i.e., the network's \emph{motif profile}) affect network persistence, and ii) how a species motif participation affects its persistence. 
    We analysed this for different levels of disturbance to the basal level. 
    To provide context, we also investigated relationships between persistence and classic global (size and connectance) and local (in-degree and trophic level) network properties.

@@

        \begin{figure}[h!]
            \centering
            \includegraphics[width=.75\textwidth]{manuscript/figures/motif_proportion_lms.pdf}
            \caption{The proportions of a network's motif profile made up by the four motifs which can appear in acyclic networks varied with the size and connectance of the network. More-connected networks tended to contain relatively more omnivory motifs and fewer apparent competition motifs. The proportions of direct competition and three-species chain motifs were less variable.}
            \label{motif_proportion_lms}
        \end{figure}

        The overall motif profile of a network was related to the average persistence of consumers ($F_{1,2998}$=571, $p$\textless0.001). 
        Average persistence was higher in networks with higher proportions of the chain, direct competition, and apparent competition motifs (Table S3, \emph{Appendix SC}). 
        Conversely, average persistence was lower in networks with higher proportions of the omnivory motif. 
        These trends were consistent across  levels of disturbance (Fig.~\ref{motif_profile_persistence}, Table S3), \emph{Appendix SC}).% Network persistence varies between networks with different dispersion of motif profiles ($F_{134,2865}$=2.19$\times10^{24}$, $p$=\textless0.001). For example, networks with more variable motif profiles have higher persistence ($\beta$=0.183, $p$\textless0.001).

        \begin{figure}
            \centering
            \includegraphics[width=.5\textwidth]{figures/persistence_motif_profiles.eps}
            \caption{The proportions of the four motifs in a network's motif profile were related to the mean persistence of species in the network. Specifically, persistence decreased as the proportion of omnivory in the network's motif profile increased while persistence increased with increasing proportions of the other three motifs. Interactions between motif profiles and disturbance were significant but small. Here we show the relationships of probabilities of extinction of basal resources of 0.1 (top set of lines) and 0.5 (bottom set of lines). Relationships are shown for the observed range of proportions for each motif.}      
            \label{fig:motif_profile_persistence}
        \end{figure}

       The relationship between frequency of participation in a motif and a species' probability of persistence depended on the level of disturbance. 
       When disturbance was low, species' persistence increased with increasing proportions of the omnivory and direct competition motifs. 
       Conversely, persistence decreased with an increasing proportion of the apparent competition motif. 
       Persistence was not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}).

        At high levels of disturbance, a species' persistence decreased as the proportion of the three-species chain increased, but persistence increased as the proportion of any other motif increased (Figure~\ref{fig:prop_lmer_all}). 
        This means that increased disturbance strengthened the effects of the three-species chain motif and changeed the direction of the effect of the apparent competition motif  (Table S4, \emph{Appendix SD}).
        The relationship between omnivory and persistence was still positive but weaker at higher levels of disturbance, and the relationship between persistence and direct competition was largely unaffected by disturbance.

            \begin{figure}[h!]
                \centering
                \includegraphics[width=\textwidth]{figures/prop_lmer_allCS.pdf}
                \caption{The effect of proportion of the role (x-axis) made up by various motifs (columns) on persistence (y-axis). The effect of participating in each motif was based on linear mixed-effect models following Equation~\ref{propreq}, except that models for each  disturbance level are fitted separately. The different colored lines indicate the probability of extinction of basal species, from $\pi_{disturbed} = 0.1$ (top, purple; no disturbance) to $\pi_{disturbed} = 0.5$ (bottom, yellow; high disturbance). Note that omnivory made up a smaller proportion of species' roles than other motifs; lines are plotted over the observed ranges of motif participation.}
                \label{fig:prop_lmer_all}
            \end{figure}

@@

        \subsubsection*{Consistency of  relationships across networks}

            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally had more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}).
            However, this trend varied at different levels of disturbance, especially for the direct competition and omnivory motifs.
            At low and medium connectances there was little change in the distribution of slopes for direct competition across all levels of disturbance. 
            At high connectances, however, the proportion of positive slopes was both lower overall and decreased with increasing disturbance. 
            For the omnivory motif these trends were reversed. 
            Moreover, the omnivory motif has a particularly broad distribution of slopes (i.e., an inconsistent relationship to persistence) when network connectance was low. 
            Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}).

        \begin{figure}[h!]
            \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_C_allS.pdf}
            \caption{Here we show the density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all sizes - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. Columns show the result for various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (left) to $\pi_{disturbed} = 0.5$ (right). Rows show various levels of connectance. The dotted, vertical line indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_C}
        \end{figure}

@@

        Species persistence varied with both in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix SF}). 
        Increased trophic level was generally associated with decreasing species persistence, although the rate of decline was slower at high disturbance (Table S5, \emph{Appendix SF}).
        In-degree showed an even stronger interaction with disturbance; having more prey species was associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

@@

            \begin{figure}[h!]
                \centering
                \includegraphics[width=.75\textwidth]{figures/roles_vs_TL.eps}
                \caption{As with motifs, persistence varied with simpler measures of species roles in networks. Moreover, motif frequencies correlated with in-degree and trophic level. \textbf{A-B)} Persistence increased with increasing in-degree for low levels of disturbance to basal resources but decreased with increasing degree for high levels of disturbance to basal resources.
                Persistence declined with inceasing STL for all levels of disturbance to basal resources.
                \textbf{C-D)} The proportion of omnivory increased with increasing degree while all other proportions decreased. The proportions of omnivory and direct competition decreased with increasing STL while the proportions of apparent competition and three-species chains increased.}
                \label{fig:motifs_vs_TL_and_deg}
            \end{figure}
